---
title: "Sanford Competition EDA - DSU"
author: "Martin Kloster"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set-Up
```{r Libraries, include=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(GGally)
library(tidyquant)
library(timeDate)
library(openxlsx)
```

```{r Data_Import}
Sanford_ED <- read.csv("../DSU-Dataset.csv")
```

# Data Wrangling
## Custom Functions
``` {r Custom_Functions}
Time_Fielding <- function(Dataset){
  Dataset %>%
    mutate(Year = as.factor(year(DateTime)),
         Month = as.factor(month(DateTime)),
         Day = as.factor(day(DateTime)),
         Shift = as.factor((hour(DateTime) %/% 6) + 1),
         Hour = as.factor(hour(DateTime)),
         .after = DateTime)
}

# Have to change this. Currently averages averages.
Timetable_Wrangling <- function(Timetable_Data, Time_Unit){
  Timetable_Data %>%
    group_by(Site, {{Time_Unit}}) %>%
    summarise(Events = sum(Events),
              across(ends_with("Sum"), sum),
              across(ends_with("Avg"), mean),
              .groups = "drop")
}

add_calendar_flags <- function(df, tz = "America/Chicago") {

  # Use the local calendar date for each DateTime
  df <- df %>%
    mutate(Date = as.Date(with_tz(DateTime, tz)))

  years <- sort(unique(lubridate::year(df$Date)))

  # --- Baseline "holiday" list: NYSE standard closures (includes observed holidays)
  nyse_holidays <- as.Date(timeDate::holidayNYSE(years, type = "standard"))

  # --- Custom “major” holidays ---
  fixed_md <- c("03-17", # St. Patrick's Day
                "05-05", # Cinco de Mayo
                "10-31", # Halloween
                "12-31", # New Year's Eve
                "12-24"  # Christmas Eve
  )

  fixed_major <- as.Date(unlist(lapply(years, function(y) paste0(y, "-", fixed_md))))

  # Traveling/family holidays
  thanksgiving <- as.Date(timeDate::holiday(years, "USThanksgivingDay"))
  black_friday <- thanksgiving + 1

  major_holidays <- sort(unique(c(
    thanksgiving, black_friday, nyse_holidays, fixed_major
  )))

  # --- Define workday and your two requested booleans ---
  df %>%
    mutate(
      # Workday = Mon–Fri and not in holiday list
      Workday = (wday(Date, week_start = 1) <= 5) & !(Date %in% nyse_holidays),
      # 1) "Last day before a workday"
      LastDayBeforeWorkday = (!Workday) & (
        # Tomorrow is a workday:
        ((wday(Date + days(1), week_start = 1) <= 5) & !((Date + days(1)) %in% nyse_holidays))
      ),
      # 2) “Major holiday”
      MajorHoliday = Date %in% major_holidays
    ) %>%
    select(-Date)  # optional: drop helper column
}
```

## Data Creation
  The original dataset has an observation for every Emergency Department (ED) Event. An Event is when people at least present at the ED, from which they may be admitted. The dataset does not include hours on site where there were no ED Events. This means that the data must first be wrangled into a format that is compatible with time series analysis (TSA).
  *Note: The original dataset wrangling will have to be cleaned up and reapplied (where appropriate) to the TSA compatible dataset. Most filtered datasets and visualizations will have to be reworked in following code.*
``` {r Data_Creation}
# ========== General Variables ==========
# Sites relevant for later crossing
Sanford_Sites <- c('A','B','C','D')
Drop_Fields <- c("Date", "Hour")

# ========== Original Dataset ==========
# Creates DateTime for future datasets
Sanford_ED <- Sanford_ED %>%
  mutate(DateTime = as.POSIXct(paste(Date, sprintf("%02d:00:00", Hour),tz="America/Chicago")), .after = Site) %>%
  select(-Date, -Hour)

# ========== Time Table ==========
# Timeframe for Timetable creation
Timeframe.Start <- as.POSIXct("2018-01-01 00:00:00", tz="America/Chicago")
Timeframe.End <- as.POSIXct("2025-08-31 23:00:00", tz="America/Chicago")

# Time Table Skeleton
Timeframe <- data.frame(
  DateTime = seq(
    from = Timeframe.Start,
    to = Timeframe.End,
    by = "hour"))

# Cross-Join Timetable, Sites required for joining ED Events and Timetable
Timetable.Sites <- crossing(Site = Sanford_Sites, DateTime = Timeframe$DateTime)

# Summarizes ED Visits for merging to Timetable
Summary.Hourly <- Sanford_ED %>%
  group_by(Site, DateTime) %>%
  summarise(Events = n(),
            Presented.Sum = sum(ED.Enc),
            Presented.Avg = mean(ED.Enc),
            Admissions.Sum = sum(ED.Enc.Admitted),
            Admissions.Avg = mean(ED.Enc.Admitted),
            .groups = "drop")

# Central timetable. Accounts for DateTimes w/o Events.
Sanford_ED.Timetable <- left_join(
  Timetable.Sites,
  Summary.Hourly,
  by = join_by(Site, DateTime)) %>%
  mutate(across(where(is.numeric), ~ replace_na(.x, 0)))

# Create Time related fields. Timetable is Time Series Analysis ready. 
Sanford_ED.Events <- Time_Fielding(Sanford_ED)
Sanford_ED.Timetable <- Time_Fielding(Sanford_ED.Timetable)
Sanford_ED.Timetable <- add_calendar_flags(Sanford_ED.Timetable)

# ========== Events Data ==========
# Summarize by Reason for Visit
Events.Reasons <- Sanford_ED.Events %>%
  group_by(REASON_VISIT_NAME) %>%
  summarise(Events = n(),
            Presentations.Sum = sum(ED.Enc),
            Presentations.Avg = mean(ED.Enc),
            Admissions.Sum = sum(ED.Enc.Admitted),
            Admissions.Avg = mean(ED.Enc.Admitted),
            Conversion_Rate = sum(ED.Enc.Admitted) / sum(ED.Enc))
```

## Data Filtering
  *Note: Will need to rework this section with the new TSA compatible dataset. Feels like this could be handled better anyways. Will have to check with professors to see if there is a better approach (IE filtering during visualization)*

``` {r Filtered_Data}
# ========== Timetables ==========
# Split into Various summary tables compatible with TSA
Timetable.Year <- Timetable_Wrangling(Sanford_ED.Timetable, Year)
Timetable.Month <- Timetable_Wrangling(Sanford_ED.Timetable, Month)
Timetable.Day <- Timetable_Wrangling(Sanford_ED.Timetable, Day)
Timetable.Shift <- Timetable_Wrangling(Sanford_ED.Timetable, Shift)
Timetable.Hour <- Timetable_Wrangling(Sanford_ED.Timetable, Hour)

# ========== Events ==========
# Events by Hour
# Events.Hourly <- Sanford_ED[c(1,2,5)] %>%
#   group_by(Site, Date, Hour) %>%
#   summarise(cases = n(), .groups = "drop") %>%
#   group_by(Site, Hour) %>%
#   summarise(cases.count = sum(cases),
#             cases.avg = mean(cases),
#             .groups = "drop")
# 
# # Events by Hour (Filtered: Oct, Nov)
# Events.Hourly.Oct_Nov <- Sanford_ED[c(1,2,5)] %>%
#   filter(month(Date) %in% c(10,11)) %>%
#   group_by(Site, Date, Hour) %>%
#   summarise(cases = n(), .groups = "drop") %>%
#   group_by(Site, Hour) %>%
#   summarise(cases.count = sum(cases),
#             cases.avg = mean(cases),
#             .groups = "drop")
# 
# # Events by Years and Hours
# Events.Hourly_Years <- Sanford_ED[c(1:3,5)] %>%
#   group_by(Site, Date, Year, Hour) %>%
#   summarise(cases = n(), .groups = "drop") %>%
#   group_by(Site, Year, Hour) %>%
#   summarise(cases.count = sum(cases),
#             cases.avg = mean(cases),
#             .groups = "drop")
# 
# # Events by Years and Seasons
# Events.Seasonal_Years <- Sanford_ED[c(1:4)] %>%
#   group_by(Site, Date, Year, Season) %>%
#   summarise(cases = n(), .groups = "drop") %>%
#   group_by(Site, Year, Season) %>%
#   summarise(cases.count = sum(cases),
#             cases.avg = mean(cases),
#             .groups = "drop")
# 
# # ========== Admissions ==========
# # Admissions by Hour
# Admissions.Hourly <- Sanford_ED %>%
#   group_by(Site, Date, Hour) %>%
#   summarise(Admissions = sum(ED.Enc.Admitted),.groups = "drop") %>%
#   group_by(Site, Hour) %>%
#   summarise(Admissions.avg = mean(Admissions), .groups = "drop")
# 
# # Admissions by Shift
# Admissions.Shift <- Sanford_ED %>%
#   group_by(Site, Date, Shift) %>%
#   summarise(Admissions = sum(ED.Enc.Admitted),.groups = "drop") %>%
#   group_by(Site, Shift) %>%
#   summarise(Admissions.avg = mean(Admissions), .groups = "drop")
```

# Data Visualization
  *Note: As mentioned earlier, none of the following visualizations mean anything substantial (or work) until it has been moved over to the TSA compatible dataset. Start with moving old mutations to the new dataset, then work on overhauling Data Filtering, and then cascade the changes down to here.*

```{r Holiday Testing}
Sanford_ED.Daily <- Sanford_ED.Timetable %>%
  mutate(Date = as.Date(DateTime)) %>%
  group_by(Site, Date, Workday, LastDayBeforeWorkday, MajorHoliday) %>%
  summarise(
    Presentations = sum(Presented.Sum),
    Admissions = sum(Admissions.Sum),
    .groups = "drop"
  )

create_flag_summary <- function(data, flag_var) {
  data %>%
    group_by({{flag_var}}) %>%
    summarise(
      Avg_Presentations = mean(Presentations),
      Avg_Admissions = mean(Admissions),
      .groups = "drop"
    )
}

Workday_Summary <- create_flag_summary(Sanford_ED.Daily, Workday)
LastDay_Summary <- create_flag_summary(Sanford_ED.Daily, LastDayBeforeWorkday)
Holiday_Summary <- create_flag_summary(Sanford_ED.Daily, MajorHoliday)


ggplot(
  Workday_Summary %>%
    pivot_longer(-Workday, names_to = "Metric", values_to = "Average"),
  aes(x = as.factor(Workday), y = Average)
) +
  geom_col(position = "dodge") +
  facet_wrap(~Metric, scales = "free_y") +
  labs(
    title = "Average Daily ER Presentations & Admissions\nWorkday vs Non-Workday",
    x = "Workday",
    y = "Average Daily Count"
  )

ggplot(
  LastDay_Summary %>%
    pivot_longer(-LastDayBeforeWorkday, names_to = "Metric", values_to = "Average"),
  aes(x = as.factor(LastDayBeforeWorkday), y = Average)
) +
  geom_col(position = "dodge") +
  facet_wrap(~Metric, scales = "free_y") +
  labs(
    title = "Average Daily ER Presentations & Admissions\nLast Day Before Workday",
    x = "Last Day Before Workday",
    y = "Average Daily Count"
  )

ggplot(
  Holiday_Summary %>%
    pivot_longer(-MajorHoliday, names_to = "Metric", values_to = "Average"),
  aes(x = as.factor(MajorHoliday), y = Average)
) +
  geom_col(position = "dodge") +
  facet_wrap(~Metric, scales = "free_y") +
  labs(
    title = "Average Daily ER Presentations & Admissions\nMajor Holiday vs Non-Holiday",
    x = "Major Holiday",
    y = "Average Daily Count"
  )

fit_single_predictor_lms <- function(data, outcome, predictor){
  data %>%
    group_by(Site) %>%
    nest() %>%
    mutate(
      model = map(data, ~ lm(reformulate(predictor, outcome), data = .x)),
      tidy  = map(model, broom::tidy),
      glance = map(model, broom::glance)
    ) %>%
    select(Site, model, tidy, glance) %>%
    unnest(tidy) %>%
    mutate(outcome = outcome, predictor = predictor)
}

# Presentations models
lm_pres_workday   <- fit_single_predictor_lms(Sanford_ED.Daily, "Presentations", "Workday")
lm_pres_lastday   <- fit_single_predictor_lms(Sanford_ED.Daily, "Presentations", "LastDayBeforeWorkday")
lm_pres_holiday   <- fit_single_predictor_lms(Sanford_ED.Daily, "Presentations", "MajorHoliday")

# Admissions models
lm_adm_workday    <- fit_single_predictor_lms(Sanford_ED.Daily, "Admissions", "Workday")
lm_adm_lastday    <- fit_single_predictor_lms(Sanford_ED.Daily, "Admissions", "LastDayBeforeWorkday")
lm_adm_holiday    <- fit_single_predictor_lms(Sanford_ED.Daily, "Admissions", "MajorHoliday")

# Combine into one table
lm_single_all <- bind_rows(
  lm_pres_workday, lm_pres_lastday, lm_pres_holiday,
  lm_adm_workday,  lm_adm_lastday,  lm_adm_holiday
)

lm_single_effects <- lm_single_all %>%
  filter(term != "(Intercept)") %>%
  arrange(outcome, predictor, Site)

lm_single_effects

fit_full_lms <- function(data, outcome){
  data %>%
    group_by(Site) %>%
    nest() %>%
    mutate(
      model = map(data, ~ lm(reformulate(c("Workday","LastDayBeforeWorkday","MajorHoliday"), outcome), data = .x)),
      tidy  = map(model, broom::tidy),
      glance = map(model, broom::glance)
    ) %>%
    select(Site, model, tidy, glance) %>%
    unnest(tidy) %>%
    mutate(outcome = outcome, model_type = "full")
}

lm_full_presentations <- fit_full_lms(Sanford_ED.Daily, "Presentations")
lm_full_admissions    <- fit_full_lms(Sanford_ED.Daily, "Admissions")

lm_full_all <- bind_rows(lm_full_presentations, lm_full_admissions)

lm_full_effects <- lm_full_all %>%
  filter(term != "(Intercept)") %>%
  arrange(outcome, Site, term)

lm_full_effects

Sanford_ED.Daily_Site2 <- Sanford_ED.Daily %>%
  mutate(
    DOW   = factor(wday(Date, label = TRUE, week_start = 1)),
    Month = factor(month(Date, label = TRUE))
  )

fit_controlled_lms <- function(data, outcome){
  data %>%
    group_by(Site) %>%
    nest() %>%
    mutate(
      model = map(data, ~ lm(
        reformulate(c("Workday","LastDayBeforeWorkday","MajorHoliday","DOW","Month"), outcome),
        data = .x
      )),
      tidy = map(model, broom::tidy),
      glance = map(model, broom::glance)
    ) %>%
    select(Site, tidy, glance) %>%
    unnest(tidy) %>%
    mutate(outcome = outcome, model_type = "controlled")
}

lm_ctrl_presentations <- fit_controlled_lms(Sanford_ED.Daily_Site2, "Presentations")
lm_ctrl_admissions    <- fit_controlled_lms(Sanford_ED.Daily_Site2, "Admissions")

lm_ctrl_effects <- bind_rows(lm_ctrl_presentations, lm_ctrl_admissions) %>%
  filter(term %in% c("WorkdayTRUE","LastDayBeforeWorkdayTRUE","MajorHolidayTRUE")) %>%
  arrange(outcome, Site, term)

lm_ctrl_effects

glance_by_site <- function(data, outcome, formula_terms){
  data %>%
    group_by(Site) %>%
    nest() %>%
    mutate(
      model = map(data, ~ lm(reformulate(formula_terms, outcome), data = .x)),
      g = map(model, broom::glance)
    ) %>%
    select(Site, g) %>%
    unnest(g) %>%
    mutate(outcome = outcome)
}

glance_full <- bind_rows(
  glance_by_site(Sanford_ED.Daily,  "Presentations", c("Workday","LastDayBeforeWorkday","MajorHoliday")),
  glance_by_site(Sanford_ED.Daily,  "Admissions",    c("Workday","LastDayBeforeWorkday","MajorHoliday"))
)

glance_full %>% select(Site, outcome, r.squared, adj.r.squared, sigma, statistic, p.value)

# Write File
write.xlsx(
  list(
    Single = lm_single_effects,
    Full = lm_full_effects,
    Controlled = lm_ctrl_effects
  ),
  file = "ED_LM_Results_By_Site.xlsx"
)
```

```{r Data_Plotting.Trends}
# Total Admissions per hour; Continuous (Not very useful)
Trend.Admissions.Sum.LC <- Sanford_ED.Timetable %>%
  filter(Year == 2024, Month %in% c(9,10)) %>%
  ggplot(aes(
    x=DateTime,
    y=Admissions.Sum,
    group = Site,
    color = Site)) +
  geom_line(linewidth = 1) +
  labs(title = "Total Admissions by Site (Over time)")
```

```{r Data_Plotting.Cycles}
# ========== Events ==========
# Top Causes of Presentations (Bar Chart)
Summary.Causes.Presentations.BC <- Sanford_ED.Events %>%
  ggplot(aes(
    x= REASON_VISIT_NAME,
    y= sum(ED.Enc))) +
  geom_bar(stat="identity") +
  coord_flip() +
  labs(title = "Top Admissions for ED Presentations (All Months)") +
  theme_bw()

# Line Chart: Events by Hour (Filtered: Oct, Nov)
# Events.Hourly.Oct_Nov.LC <- Events.Hourly.Oct_Nov %>%
#   ggplot(aes(
#     x=Hour,
#     y=cases.avg,
#     group = Site,
#     color = Site)) +
#   geom_line(linewidth = 1) +
#   labs(title = "Hourly Average Visits by Site: October, November") +
#   theme_bw()

# Line Chart: Events by Hour (Faceted: Years)
#Events.Hourly.Facet_Years.LC <- Events.Hourly_Years %>%
#  ggplot(aes(
#    x=Hour,
#    y=cases.count,
#    group = Site,
#    color = Site)) +
#  geom_line(linewidth = 1) +
#  facet_wrap(~Year) +
#  labs(title = "Hourly Average Visits by Site: All Months") +
#  theme_bw()

# Line Chart: Events by Season (Facet: Years)
# Events.Seasonal.Facet_Years.LC <- Events.Seasonal_Years %>%
#   ggplot(aes(
#     x=Season,
#     y=cases.count,
#     group = Site,
#     color = Site)) +
#   geom_line(linewidth = 1) +
#   facet_wrap(~Year) +
#   labs(title = "Hourly Average Visits by Site: All Months") +
#   theme_bw()

# ========== Admissions ==========
# Avg ED Admissions by Hour (Line Chart)
Cycle.Admissions.Avg.Hourly.LC <- Timetable.Hour %>%
  ggplot(aes(
    x=Hour,
    y=Admissions.Avg,
    group = Site,
    color = Site)) +
  geom_line(linewidth = 1) +
  labs(title = "Average Admissions by Site, Hour (All Months)") +
  theme_bw()

# Avg ED Admissions by Shift (Boxplot)
Cycle.Admissions.Shift.BP <- Sanford_ED.Timetable %>%
  ggplot(aes(
    x=Shift,
    y=Admissions.Sum)) +
  geom_boxplot() +
  facet_wrap(~Site) +
  labs(title = "Admissions by Site, Shift (All Months)") +
  theme_bw()

# Line Chart: ED Admissions by Hour (Obsolete)
#Admissions.Hourly.LC <- Admissions.Hourly %>%
#  ggplot(aes(
#    x=Hour,
#    y=Admissions.avg,
#    group = Site,
#    color = Site)) +
#  geom_line(linewidth = 1) +
#  labs(title = "Avg Admissions by Hour, Site (All Months)") +
#  theme_bw()

# Line Chart: ED Admissions by Shift (Obsolete)
#Admissions.Shift.LC <- Admissions.Shift %>%
#  group_by(Shift) %>%
#  ggplot(aes(
#    x=Shift,
#    y=Admissions.avg,
#    group = Site,
#    color = Site)) +
#  geom_line(linewidth = 1) +
#  labs(title = "Avg Admissions by Shift, Site (All Months)") +
#  theme_bw()
```

``` {r Display}
# ========== Trends ==========
Trend.Admissions.Sum.LC    # Too messy to be usable. Keep Looking for Modifications

# ========== Cycles ==========
# Events
#Events.Hourly.Facet_Years.LC
#Events.Hourly.Oct_Nov.LC
#Events.Seasonal.Facet_Years.LC

# Number Presenting to ED

# ED Admissions
Cycle.Admissions.Avg.Hourly.LC
Cycle.Admissions.Shift.BP
#Admissions.Hourly.LC
#Admissions.Shift.LC

Summary.Causes.Presentations.BC
```
